Measuring neutron star tidal deformability with Advanced LIGO: a Bayesian analysis of neutron star - black hole binary observations

### 2016-09-06 Paper/paper.tex
shrunk fig 9

\label{fig:}

@@

In the previous section, we showed that the effects of the deformation of NSs
due to the quadrupolar tidal field of their companion BHs become discernible in
the GW spectrum under certain favorable conditions, including (a) BH mass is
sufficiently small, (b) BH spin is positive aligned, i.e. $\chibh\gtrsim +0.4$,
(c) the NS is not very compact, with $\lambdans\gtrsim 1000$, and (d) the
source location and orientation are such that its GW SNR $\gtrsim 30$.
% 
Both condition (a) and (b) enhance the tidal distortion of the star and increase
the number of orbits the system goes through at small separation, where the
differences between NSBH and BBH signals are maximal.
% 
Conditions (a)-(c) also reduce the onset frequency of the disruption of the NS,
allowing for it to happen earlier in the orbit. 
% 
We expect that these conditions are also the ones which would maximize the
likelihood of {\it measuring} tidal effects in NSBH signals. In this section,
%
%In addition, a SNR of $30$ corresponds to a $1.35+5M_\odot$ binary with 
%$\chibh = 0.4$ at a luminosity distance of $XXX$~Mpc, if optimally aligned
%with LIGO ($XXX$~Mpc if the binary subtends a $45^o$ angle with LIGO's line of
%sight.
%
%These conditions are strong, but not unrealistic. For e.g., a detection rate
%of $20$ NSBHs a year implies $1-2$ of them will have SNRs$\gtrsim 30$, since
%the number of expected events scales as $1/\rho^2$ if we assume that sources
%are distributed uniformly in spacial volume.
%
we turn the question around and ask: under similarly favorable circumstances, can
we gain insights about the internal structure of neutron stars from GW observations.

In this section, we calculate the accuracy with which we measure $\lambdans$ from
{\it single} GW observations. We sample the same set of disruptive NSBH mergers
as in the previous section, i.e. those with $q=\{2,3,4,5\}$,
$\chibh=\{-0.5,0,+0.5,+0.75\}$, and $\lambdans=\{500, 800, 1000, 1500, 2000\}$;
fixing the NS mass $\mns=1.35 M_\odot$ and $\chins=0$. For each unique
combination of these parameters, we inject LEA+ signals into zero noise and
perform a fully Bayesian parameter estimation analysis of each with LEA+
templates. Our priors on component masses and spins remain as in the previous
section, with mass-ratio additionally restricted to $2\leq q\leq 6$, and
$\lambdans$ sampled uniformly from $[0,4000]$.
% 
% 
%
As an illustration, we show the recovered probability distribution for
$\lambdans$ for three specific configurations in 
Fig.~\ref{fig:SingleSystemLambdaPDFvsSNR}. We fix
$q = \mbh /\mns = 5.4M_\odot/1.35M_\odot = 4$, with $\chibh=+0.5$, while
varying $\lambdans$ over $\{1000, 1500, 2000\}$ between the three panels.
The SNR is fixed at $\rho=50$. Both the true and measured values of
$\lambdans$ are marked. The darker shaded region marks the $90\%$ credible
interval on $\lambdans$. We note that $\lambdans$ is estimated to within
$\pm 2000$ of its true value at this SNR. Another interesting thing to note 
is that while $(\Delta\lambdans)^{90\%}$ slowly grows with $\lambdans$, the
fractional uncertainty
\begin{equation}
\delta\lambdans^{90\%}:= (\Delta\lambdans)^{90\%}/\lambdans
\end{equation}
decreases instead.
%
\textcolor{blue}{%
Further illustrations of the measurements of mass, spin and tidal parameters,
as well as their correlations, are shown in Appendix~\ref{as1:illustrations}.
}
% 
\textcolor{gray}{%
Next, in Fig.~\ref{fig:SingleSystemLambda2DPDFs} we show the correlation of
mass, spin, and tidal parameter measurements. We keep the binary parameters
as in Fig.~\ref{fig:SingleSystemLambdaPDFvsSNR}, with $\lambdans=2000$, and
set $\rho=30$ (left panel) or $\rho=50$ (right panel).
%
We find that the measurement of $\lambdans$ is weakly degenerate with
other parameters, and at realistic SNRs it would improve by a few tens of 
percent if we knew non-tidal parameters to better accuracy. The predominant
factor that would enhance the measurement accuracy for $\lambdans$ is the
signal strength, and only when $\rho\gtrsim 50$ do we expect $\lambdans$ 
measurement to be limited by its degeneracy with non-tidal parameters at 
a factor of few level, as was reported by Ref.~\cite{Lackey:2013axa}.
}

In Fig.~\ref{fig:TT_LambdaCIWidths90_0_Lambda_SNR} we show the main results of
this section. In each panel, as a function of black hole mass and spin, we show
the measured $90\%$ credible interval widths $(\Delta\lambdans)^{90\%}$. These
correspond to the full width of the dark shaded regions in the illustrative
Fig.~\ref{fig:SingleSystemLambdaPDFvsSNR}. The effect of increasing signal
strength can be seen as we go from left to right in each row. The effect of the
NS tidal deformability parameter $\lambdans$ on its own measurability can be
seen by comparing panels within each column, with the NS becoming more
deformable from top to bottom. 
%
A uniform pattern emerges in the left-most column, which corresponds to $\rho=20$.
We find that at this signal strength, our measurement of $\lambdans$ is
dominated by the width of our prior on it. The $90\%$ credible intervals span
the entire allowed range for $\lambdans$, making a reasonable estimation of
$\lambdans$ at $\rho\simeq20$ difficult.
% 
Increasing the signal strength to $\rho=30$ gives marginally better results,
bringing down the statistical uncertainties to within $\pm 75-100\%$ of the
true $\lambdans$ value~\footnote{The symmetric error-bars of $\pm\mathrm{X}\%$
correspond to $\dlambda = 2\mathrm{X}\%$.}.
%
It is not until we reach an SNR as high as $\rho\gtrsim 50$, can we put
meaningful (i.e. $\mathcal{O}(10\%)$) constraints on $\lambdans$. For e.g.,
with a {\it single} observation of a $q=4$ binary with $\chibh\geq 0.6$ and
$\rho = 50$~\footnote{For an optimally oriented source with
$q=4, \mns=1.35M_\odot, \chibh=0.6$, an SNR of $\rho = 50$ corresponds to
a luminosity distance of $\approx 113$Mpc.}, we would be able to estimate 
$\lambdans$ to within $\pm 40\%$ of its true value (which is equivalent to
measuring the ratio of NS radius to mass with an uncertainty of about
$\pm 10\%$).
% 
The results presented so far are quite in agreement with Sec.~\ref{s1:PEwithnoNS}.

Amongst other source parameters, BH mass and spin play a dominant role. A smaller
BH with a larger spin always allows for a more precise measurement on $\lambdans$.
Visually we can see this in the bottom right corner of each panel in
Fig.~\ref{fig:TT_LambdaCIWidths90_0_Lambda_SNR}, which corresponds to low-mass BHs
with large spins, and is simultaneously the region of smallest measurement errors on $\lambdans$.
The actual deformability of the NS also plays an important role on its own
measurability. For e.g., when $\lambdans\leq 1000$, it is fairly difficult
to meaningfully constrain $\lambdans$ without requiring the source to be
close ($\approx 100$Mpc) with a GW SNR $\rho\gtrsim 50$. Quantifying this further,
in Fig.~\ref{fig:TT_SNRThresholds_BHspin_BHmass_CI90_0} we show the minimum
signal strength required to attain a certain level of credibility in our
$\lambdans$ measurement, as a function of BH properties. The NS is allowed
the most favorable (hardest) EoS considered, with $\lambdans^\mathrm{true}=2000$.
% 
We first note that, even with the most favorable BH and NS properties, achieving
a $\pm 50\%$ measurement certainty on $\lambdans$ will require a GW SNR
$\rho\gtrsim 30$. If we additionally restrict BH masses to lie outside of the so-called
astrophysical mass-gap~\cite{Bailyn:1997xt,Kalogera:1996ci,Kreidberg:2012,
Littenberg:2015tpa}, we will simultaneously need to restrict BH spins
to $\chibh\gtrsim +0.5$ to obtain the same measurement credibility at the same
source location.

@@

\textbf{Improvement in $\lambdans$ measurement accuracy with multiple observations}\par\medskip
\includegraphics[width=1.05\columnwidth,trim=1cm 0 0 0]{plots/pdfLambda_vs_N_L800.pdf}
\includegraphics[width=1.\columnwidth,trim=0 0 1cm 0]{plots/FillBetweenErrorBarsLambda_vs_N_L800.pdf}

>>>>

\label{fig:TT_SNRThresholds_BHspin_BHmass_CI90_0}

@@

In the previous section, we showed that the effects of the tidal deformation of
NSs by their companion BHs become discernible in
the GW spectrum under certain favorable conditions, including (a) BH mass is
sufficiently small, (b) BH spin is positive aligned, i.e. $\chibh\gtrsim +0.4$,
(c) the NS is not very compact, with $\lambdans\gtrsim 1000$, and (d) the
source location and orientation are such that its GW SNR $\gtrsim 30$.
% 
Both condition (a) and (b) enhance the tidal distortion of the star and increase
the number of orbits the system goes through at small separation, where the
differences between NSBH and BBH signals are maximal.
% 
Conditions (a)-(c) also reduce the onset frequency of the disruption of the NS,
allowing for it to happen earlier in the orbit. 
% 
We expect that these conditions are also the ones which should maximize the
likelihood of {\it measuring} tidal effects in NSBH signals. In this section,
%
%In addition, a SNR of $30$ corresponds to a $1.35+5M_\odot$ binary with 
%$\chibh = 0.4$ at a luminosity distance of $XXX$~Mpc, if optimally aligned
%with LIGO ($XXX$~Mpc if the binary subtends a $45^o$ angle with LIGO's line of
%sight.
%
%These conditions are strong, but not unrealistic. For e.g., a detection rate
%of $20$ NSBHs a year implies $1-2$ of them will have SNRs$\gtrsim 30$, since
%the number of expected events scales as $1/\rho^2$ if we assume that sources
%are distributed uniformly in spacial volume.
%
we turn the question around to ask: under similarly favorable circumstances,
can we gain insights about the internal structure of neutron stars from GW
observations?

In this section, we calculate the accuracy with which we measure $\lambdans$ from
{\it single} GW observations. We sample the same set of disruptive NSBH mergers
as in the previous section, i.e. those with $q=\{2,3,4,5\}$,
$\chibh=\{-0.5,0,+0.5,+0.75\}$, and $\lambdans=\{500, 800, 1000, 1500, 2000\}$;
fixing the NS mass $\mns=1.35 M_\odot$ and $\chins=0$. For each unique
combination of these parameters, we inject LEA+ signals into zero noise and
perform a fully Bayesian parameter estimation analysis of each with LEA+
templates. Our  priors on component masses and spins remain as in the
previous section, with mass-ratio additionally restricted to $2\leq q\leq 6$,
and $\lambdans$ sampled uniformly from $[0,4000]$.
% 
%
As an illustration of individual injections, we show the recovered probability
distribution for $\lambdans$ for three specific configurations in 
Fig.~\ref{fig:SingleSystemLambdaPDFvsSNR}. We fix
$q = \mbh /\mns = 5.4M_\odot/1.35M_\odot = 4$, with $\chibh=+0.5$, and
vary $\lambdans$ over $\{1000, 1500, 2000\}$ between the three panels.
The SNR is fixed at $\rho=50$. The darker shaded regions mark the $90\%$ credible
interval on $\lambdans$. We note that $\lambdans$ is estimated to within
$\pm 2000$ of its true value at this SNR. Another interesting thing to note 
is that while $(\Delta\lambdans)^{90\%}$ slowly grows with $\lambdans$, the
fractional uncertainty
\begin{equation}
\delta\lambdans^{90\%}:= (\Delta\lambdans)^{90\%}/\lambdans
\end{equation}
decreases instead.
%
\textcolor{blue}{%
Further illustrations of the measurements of non-tidal parameters,
as well as their correlations, are shown in Appendix~\ref{as1:illustrations}.
We will continue here to focus on the measurement of $\lambdans$ itself.
}
% 
\textcolor{gray}{%
Next, in Fig.~\ref{fig:SingleSystemLambda2DPDFs} we show the correlation of
mass, spin, and tidal parameter measurements. We keep the binary parameters
as in Fig.~\ref{fig:SingleSystemLambdaPDFvsSNR}, with $\lambdans=2000$, and
set $\rho=30$ (left panel) or $\rho=50$ (right panel).
%
We find that the measurement of $\lambdans$ is weakly degenerate with
other parameters, and at realistic SNRs it would improve by a few tens of 
percent if we knew non-tidal parameters to better accuracy. The predominant
factor that would enhance the measurement accuracy for $\lambdans$ is the
signal strength, and only when $\rho\gtrsim 50$ do we expect $\lambdans$ 
measurement to be limited by its degeneracy with non-tidal parameters at 
a factor of few level, as was reported by Ref.~\cite{Lackey:2013axa}.
}

In Fig.~\ref{fig:TT_LambdaCIWidths90_0_Lambda_SNR} we show the main results of
this section. In each panel, as a function of black hole mass and spin, we show
the measured $90\%$ credible interval widths $(\Delta\lambdans)^{90\%}$. These
correspond to the full width of the dark shaded regions in the illustrative
Fig.~\ref{fig:SingleSystemLambdaPDFvsSNR}. The effect of increasing signal
strength can be seen as we go from left to right in each row. The effect of the
NS tidal deformability parameter $\lambdans$ on its own measurability can be
seen by comparing panels within each column, with the NS becoming more
deformable from top to bottom. 
%
A uniform pattern emerges in the left-most column, which corresponds to $\rho=20$.
We find that at this signal strength, our measurement of $\lambdans$ is
dominated by the width of our prior on it. The $90\%$ credible intervals span
the entire allowed range for $\lambdans$, making a reasonable estimation of
$\lambdans$ at $\rho\simeq20$ difficult.
% 
Increasing the signal strength to $\rho=30$ gives marginally better results,
bringing down the statistical uncertainties to within $\pm 75-100\%$ of the
true $\lambdans$ value~\footnote{The symmetric error-bars of $\pm\mathrm{X}\%$
correspond to $\dlambda = 2\mathrm{X}\%$.}.
%
It is not until we reach an SNR as high as $\rho\gtrsim 50$, can we put
meaningful (i.e. $\mathcal{O}(10\%)$) constraints on $\lambdans$. For e.g.,
with a {\it single} observation of a $q=4$ binary with $\chibh\geq 0.6$ and
$\rho = 50$~\footnote{For an optimally oriented source with
$q=4, \mns=1.35M_\odot, \chibh=0.6$, an SNR of $\rho = 50$ corresponds to
a luminosity distance of $\approx 113$Mpc.}, we would be able to estimate 
$\lambdans$ to within $\pm 40\%$ of its true value (which is equivalent to
measuring the ratio of NS radius to mass with an uncertainty of about
$\pm 10\%$).
% 
The results presented so far are in agreement with Sec.~\ref{s1:PEwithnoNS}.

Amongst other source parameters, BH mass and spin play a dominant role. A smaller
BH with a larger spin always allows for a more precise measurement on $\lambdans$.
We can see this in the bottom right corner of each panel in
Fig.~\ref{fig:TT_LambdaCIWidths90_0_Lambda_SNR}, which corresponds to low-mass BHs
with large spins, and is simultaneously the region of smallest measurement errors on $\lambdans$.
The actual deformability of the NS also plays an important role on its own
measurability. For e.g., when $\lambdans\leq 1000$, it is fairly difficult
to meaningfully constrain $\lambdans$ without requiring the source to be
close ($\approx 100$Mpc) with a GW SNR $\rho\gtrsim 50$. Quantifying this further,
in Fig.~\ref{fig:TT_SNRThresholds_BHspin_BHmass_CI90_0} we show the minimum
signal strength required to attain a certain level of credibility in our
$\lambdans$ measurement, as a function of BH properties. The NS is allowed
the most favorable (hardest) EoS considered, with $\lambdans^\mathrm{true}=2000$.
% 
We first note that, even with the most favorable BH and NS properties, achieving
a $\pm 50\%$ measurement certainty on $\lambdans$ will require a GW SNR
$\rho\gtrsim 30$. If we additionally restrict BH masses to lie outside of the so-called
astrophysical mass-gap~\cite{Bailyn:1997xt,Kalogera:1996ci,Kreidberg:2012,
Littenberg:2015tpa}, we will simultaneously need to restrict BH spins
to $\chibh\gtrsim +0.5$ to obtain the same measurement credibility at the same
source location.

@@

\textbf{Improvement in $\lambdans$ measurement accuracy\\ with multiple observations}\par\medskip
\includegraphics[width=0.8\columnwidth]{plots/pdfLambda_vs_N_L800.pdf}\\
\includegraphics[width=0.8\columnwidth]{plots/FillBetweenErrorBarsLambda_vs_N_L800.pdf}
